Developing an automated iterative near-term forecasting system for an ecological

### 2018-07-03 manuscript/portal_iterative_forecasting_ms.Rmd
Polishing edits to revised manuscript

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of how well models capture the processes governing ecological systems [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-@clark2001] called for a more central role of forecasting in ecology. Since then, an increasing number of ecological forecasts are being published that focus on a societally important questions from daily to decadal time scales [@dietze2018]. At daily scales, ecological forecasts predict the occurance of environmental issues like toxic algal blooms [@stumpf2009]. At quarterly scales, forecasts are used to predict the stocks of fisheries and make decisions about quotas for fishing [@NOAA2016]. At decadal time scales, ecological forecasts are used to predict how biodiversity will change as it responds to anthropogenic influences [@harris2018]. These forecasting examples highlight the important role that ecological forecasts play in recasting ecological knowledge in societally relevant ways and also improve our understanding of ecological systems by testing the ability of our models to predict how systems will change in the future [@dietze2018; @harris2018].

While some of the examples given above (e.g., fisheries stock estimates) are regularly repeated, most ecological forecasts are made once, published, and never assessed or updated [@dietze2018]. This lack of both regular assessment and active updating has limited the progress of ecological forecasting and hindered our ability to make useful and reliable predictions. The lack of active assessment results in limited information on how much confidence to place in forecasts and makes it difficult to determine on which forecasting methods to build. Without regular updates, forecasts lack the most current data, and the longer a forecast remains out of date, the less accurate it becomes [@petchey2015; @dietze2018]. More regular updating and assessment will advance ecological forecasting as a field by accelerating the identification of the best models for individual forecasts and improving our understanding of how to best design forecasting approaches for ecology in general. This approach has helped accelerate forecasting ability in other fields including meteorology [@kalnay2003; @mcgill2012; @bauer2015]. For ecological forecasting to mature as a field, we need to change how we produce and interact with forecasts, creating a more dynamic interplay between model development, prediction generation, and incorporation of new data and information [@dietze2018].

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when the `tscount` package [@liboschik2015], used by two of our forecasting models, was temporarily removed from CRAN (the R package repository) and could not be installed in the usual way. This broke our forecasting pipeline, as we could no longer run models that used that package. To make our pipeline robust to changes in external software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything needed to run a given piece of software, including the operating system. Once created, a software container is basically a time capsule, containing all the software dependencies in the exact state used to develop and run the software. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in the container. We use an existing platform, Docker [@merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to control transitions to new package versions, implementing them only after we have tested them and made any necessary changes to the data processing and analysis code. We use a container created by the Rocker project, which is a Docker image with many important R packages [i.e., the tidyverse packages; @tidyverse] pre-installed  [@boettiger2017]. We add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

Iterative forecasting benefits from frequently updated data so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Both frequent data collection and rapid processing are important for providing timely forecasts. Since we collect data monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database (Figure 1).

@@

Once data are available, it must be processed into a form appropriate for modeling (Figure 1). For many ecological datasets, this requires not only simple data manipulation but also a good understanding of the data to facilitate appropriate aggregation. Data manipulation steps are often conducted using custom one-off code to convert the raw data into the desired form [@morris2013], but this approach has several limitations. First, each researcher must develop and maintain their own data manipulation code, which is inefficient and can result in different researchers producing different versions of the data for the same task. Subtle differences in data processing decisions have led to confusion when reproducing results for the Portal data in the past. Second, this kind of code is rarely robust to changes in data structure and location. Based on our experience developing and maintaining the Data Retriever [@morris2013; @senyondo2017], these kinds of changes are common. Finally, this kind of code is generally poorly tested, which can lead to errors based on mistakes in data manipulation. To avoid these issues for the Portal Project data, the Portal team has been developing an R package (portalr; http://github.com/weecology/portalr) for acquiring the data and handling common data cleaning and aggregation tasks. As a result, our modeling and forecasting code only needs to install this package and run the data manipulation and summary functions to get the appropriate data (Figure 1b). The package undergoes thorough automated unit testing to ensure that data manipulations are achieving the desired results. Having data manipulation code maintained in a separate package that focuses on consistently providing properly summarized forms of the most recent data has made maintaining the forecasting code itself much more straightforward.

@@

Publicly archiving forecasts before new data is collected allows the field to assess, compare, and build on forecasts made by different groups [@mcgill2012; @tredennick2016; @dietze2018; @harris2018] (Figure 1). Archiving serves as a form of pre-registration for model predictions because the forecasts cannot be modified once the data to assess them has been collected. This helps facilitate an unbiased interpretation of model performance. To serve this role, archives should be publicly accessible and be a permanent record that cannot be changed or deleted. This second criterion means that GitHub is not sufficient for archival purposes because repositories can be changed or deleted [@bergman2012; @white2015].  We explored three major repositories for archiving forecasts: FigShare (https://figshare.com/), Zenodo (https://zenodo.org/), and Open Science Framework (https://osf.io/). While all three repositories allowed for easy manual submissions (i.e., a human uploading files after each forecast), automating this process was substantially more difficult. Various combinations of repositories, APIs (i.e., interfaces for automatically interacting with the archiving websites), and associated R packages had issues with: 1) integrating authorization with continuous integration; 2) automatically making archived files public; 3) adding new files to an existing location; or 4) automatically permanently archiving the files. Our eventual solution was to leverage the GitHub-Zenodo integration (https://guides.github.com/activities/citable-code/) and automatically push forecasts to a GitHub repository from the CI server and release them via the GitHub API. The GitHub-Zenodo integration is designed to automatically create versioned archives of GitHub repositories. We created a repository for storing forecasts (https://github.com/weecology/forecasts) and linked this repository with Zenodo (a one-time manual process). Each time a new forecast is created, our pipeline adds the new forecasts to the GitHub repository and uses the GitHub API to create a new "release" for that repository. This triggers the GitHub-Zenodo integration, which automatically archives the resulting forecasts under a top-level DOI that refers to all archived forecasts ([https://doi.org/10.5281/zenodo.839580](https://doi.org/10.5281/zenodo.839580)). Through this process, we automatically archive every forecast made with a documented time-stamp. In addition, we also archive the full state of the modeling and forecasting repository ([https://doi.org/10.5281/zenodo.833438](https://doi.org/10.5281/zenodo.833438)). This ensures that every forecast is fully reproducible since the exact code used to generate every forecast is preserved. Early forecasts from this system are archived in the modeling and forecasting code archive, not in the newer repository 'forecasts'.

@@

Because of the breadth of expertise needed to set up our forecasting pipeline, our effort required a team with diverse skills and perspectives, ranging from software development to field site expertise. It is rare to find such breadth within a single individual, and our system was developed as a collaboration between the lab collecting the data and a computational ecology lab. When teams have a breadth of expertise, communication can be challenging [@winowiecki2011]. We found a shared base of knowledge related to both the field research and computational skills was important for the success of the group. The two labs are part of a joint interdisciplinary ecology group that has a mission of breaking down barriers between field and computational/theoretical ecologists (http://weecology.org). Everyone on the team had received training in fundamental data management and computing skills through a combination of university courses, Software and Data Carpentry workshops [@teal2015], and lab training efforts. In addition, everyone was broadly familiar with the study site and methods of data collection, and most team members had participated in field work at the site on multiple occasions. This provided a shared set of knowledge and vocabulary that actively facilitated interdisciplinary interactions. All members of the team actively participated in the development of the forecasting pipeline. Given the current state of tools for forecasting, forecasting teams require some experience in working with continuous integration and APIs. This means either interdisciplinary teams or additional training will often be required for creating these pipelines until tool development improves. To improve the success of these diverse groups, we believe efforts at providing 'team science' training to scientists interested in forecasting will be beneficial for the success of iterative forecasting attempts for the foreseeable future [@read2016].

>>>>

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of how well models capture the processes governing ecological systems [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-@clark2001] called for a more central role of forecasting in ecology. Since then, an increasing number of ecological forecasts are being published that focus on societally important questions from daily to decadal time scales [@dietze2018]. At daily scales, ecological forecasts predict the occurance of environmental issues like toxic algal blooms [@stumpf2009]. At quarterly scales, forecasts are used to predict the stocks of fisheries and make decisions about quotas for fishing [@NOAA2016]. At decadal time scales, ecological forecasts are used to predict how biodiversity will change as it responds to anthropogenic influences [@harris2018]. These forecasting examples highlight the important role that ecological forecasts play in recasting ecological knowledge in societally relevant ways and also improve our understanding of ecological systems by testing the ability of our models to predict how systems will change in the future [@dietze2018; @harris2018].

While some of the examples given above (e.g., fisheries stock estimates) are regularly repeated, most ecological forecasts are made once, published, and never assessed or updated [@dietze2018]. This lack of both regular assessment and active updating has limited the progress of ecological forecasting and hindered our ability to make useful and reliable predictions. The lack of active assessment results in limited information on how much confidence to place in forecasts and makes it difficult to determine on which forecasting methods to build. Without regular updates, forecasts lack the most current data, and the longer a forecast remains out of date, the less accurate it becomes [@petchey2015; @dietze2018]. More regular updating and assessment will advance ecological forecasting as a field by accelerating the identification of the best models for individual forecasts and improving our understanding of how to best design forecasting approaches for ecology in general. This approach has helped accelerate forecasting ability in other fields such as meteorology [@kalnay2003; @mcgill2012; @bauer2015]. For ecological forecasting to mature as a field, we need to change how we produce and interact with forecasts, creating a more dynamic interplay between model development, prediction generation, and incorporation of new data and information [@dietze2018].

@@

In addition to automatically running software pipelines, the other key component of "continuous analysis" is making sure that the pipelines will continue to run even as software dependencies change [@boettiger2015; @beaulieu2017]. Many of us have experienced the frustrations that can occur when software updates (e.g., changes in R package versions) create errors in previously functional code. We experienced this issue when the `tscount` package [@liboschik2015], used by two of our forecasting models, was temporarily removed from CRAN (the R package repository) and could not be installed in the usual way. This broke our forecasting pipeline, as we could no longer run models that used that package. To make our pipeline robust to changes in external software dependencies, we follow Beaulieu and Greene's [-@beaulieu2017] recommendation to use software containers. Software containers are standalone packages that contain copies of everything needed to run a given piece of software, including the operating system. Once created, a software container is basically a time capsule, containing all the software dependencies in the exact state used to develop and run the software. If those dependencies change (or disappear) in the wider world, they still exist, unchanged, in the container. We use an existing platform, Docker [@merkel2014], to store an exact image of the complete software environment for running the forecasts. Docker also allows a specified set of packages to be used consistently across different computer and server environments. Using containers allows us to control transitions to new package versions, implementing them only after we have tested them and made any necessary changes to the data processing and analysis code. We use a container created by the Rocker project, which is a Docker image with many important R packages [i.e., the tidyverse packages; @tidyverse] pre-installed  [@boettiger2017]. We add our code and dependencies to this existing Rocker image to create a software container for our forecasting pipeline. In combination, the automated running of the pipeline (continuous integration) and the guarantee it will not stop working unexpectedly due to software dependencies (via a software container) allows continuous analysis to serve as the glue that connects all stages of the forecasting pipeline.

@@

Iterative forecasting benefits from frequently updated data so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Both frequent data collection and rapid processing are important for providing timely forecasts. Since we collect data monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database (@yenni2018; Figure 1).

@@

Once data are available, they must be processed into a form appropriate for modeling (Figure 1). For many ecological datasets, this requires not only simple data manipulation but also a good understanding of the data to facilitate appropriate aggregation. Data manipulation steps are often conducted using custom one-off code to convert the raw data into the desired form [@morris2013], but this approach has several limitations. First, each researcher must develop and maintain their own data manipulation code, which is inefficient and can result in different researchers producing different versions of the data for the same task. Subtle differences in data processing decisions have led to confusion when reproducing results for the Portal data in the past. Second, this kind of code is rarely robust to changes in data structure and location. Based on our experience developing and maintaining the Data Retriever [@morris2013; @senyondo2017], these kinds of changes are common. Finally, this kind of code is generally poorly tested, which can lead to errors based on mistakes in data manipulation. To avoid these issues for the Portal Project data, the Portal team has been developing an R package (portalr; http://github.com/weecology/portalr) for acquiring the data and handling common data cleaning and aggregation tasks. As a result, our modeling and forecasting code only needs to install this package and run the data manipulation and summary functions to get the appropriate data (Figure 1b). The package undergoes thorough automated unit testing to ensure that data manipulations are achieving the desired results. Having data manipulation code maintained in a separate package that focuses on consistently providing properly summarized forms of the most recent data has made maintaining the forecasting code itself much more straightforward.

@@

Publicly archiving forecasts before new data is collected allows the field to assess, compare, and build on forecasts made by different groups [@mcgill2012; @tredennick2016; @dietze2018; @harris2018] (Figure 1). Archiving serves as a form of pre-registration for model predictions because the forecasts cannot be modified once the data to assess them has been collected. This helps facilitate an unbiased interpretation of model performance. To serve this role, archives should be publicly accessible and be a permanent record that cannot be changed or deleted. This second criterion means that GitHub is not sufficient for archival purposes because repositories can be changed or deleted [@bergman2012; @white2015].  We explored three major repositories for archiving forecasts: FigShare (https://figshare.com/), Zenodo (https://zenodo.org/), and Open Science Framework (https://osf.io/). While all three repositories allowed for easy manual submissions (i.e., a human uploading files after each forecast), automating this process was substantially more difficult. Various combinations of repositories, APIs (i.e., interfaces for automatically interacting with the archiving websites), and associated R packages had issues with: 1) integrating authorization with continuous integration; 2) automatically making archived files public; 3) adding new files to an existing location; or 4) automatically permanently archiving the files. Our eventual solution was to leverage the GitHub-Zenodo integration (https://guides.github.com/activities/citable-code/) and automatically push forecasts to a GitHub repository from the CI server and release them via the GitHub API. The GitHub-Zenodo integration is designed to automatically create versioned archives of GitHub repositories. We created a repository for storing forecasts (https://github.com/weecology/forecasts) and linked this repository with Zenodo (a one-time manual process). Each time a new forecast is created, our pipeline adds the new forecasts to the GitHub repository and uses the GitHub API to create a new "release" for that repository. This triggers the GitHub-Zenodo integration, which automatically archives the resulting forecasts under a top-level DOI that refers to all archived forecasts ([https://doi.org/10.5281/zenodo.839580](https://doi.org/10.5281/zenodo.839580)). Through this process, we automatically archive every forecast made with a documented time-stamp. In addition, we also archive the full state of the modeling and forecasting repository ([https://doi.org/10.5281/zenodo.833438](https://doi.org/10.5281/zenodo.833438)). Through a similar process, the raw data in the data repository is also archived on a Zenodo whenever data is added or changed [@yenni2018], allowing retrieval of older versions of the data used for forecasting ([https://doi.org/10.5281/zenodo.1219752](https://doi.org/10.5281/zenodo.1219752)). This ensures that every forecast is fully reproducible since the exact code and data used to generate every forecast is preserved. Early forecasts from this system are archived in the modeling and forecasting code archive, not in the newer repository 'forecasts'.

@@

Because of the breadth of expertise needed to set up our forecasting pipeline, our effort required a team with diverse skills and perspectives, ranging from software development to field site expertise. It is rare to find such breadth within a single individual, and our system was developed as a collaboration between the lab collecting and managing the data and a computational ecology lab. When teams have a breadth of expertise, communication can be challenging [@winowiecki2011]. We found a shared base of knowledge related to both the field research and computational skills was important for the success of the group. The two labs are part of a joint interdisciplinary ecology group that has a mission of breaking down barriers between field and computational/theoretical ecologists (http://weecology.org). Everyone on the team had received training in fundamental data management and computing skills through a combination of university courses, Software and Data Carpentry workshops [@teal2015], and lab training efforts. In addition, everyone was broadly familiar with the study site and methods of data collection, and most team members had participated in field work at the site on multiple occasions. This provided a shared set of knowledge and vocabulary that actively facilitated interdisciplinary interactions. All members of the team actively participated in the development of the forecasting pipeline. Given the current state of tools for forecasting, forecasting teams require some experience in working with continuous integration and APIs. This means either interdisciplinary teams or additional training will often be required for creating these pipelines until tool development improves. To improve the success of these diverse groups, we believe efforts at providing 'team science' training to scientists interested in forecasting will be beneficial for the success of iterative forecasting attempts for the foreseeable future [@read2016].
